Article Title

### 2013-08-14 sections/Theory.tex
edited the Theory section

\subsection{Overview}
Given the sample $x_1,\ldots,x_n$ of $p\times 1$ observations, the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-d} \sum\limits_{i=1}^n (x_i - \bar x) (x_i - \bar x)^\T
\quad\mbox{where}\;
\bar x = \frac 1 n \sum\limits_{i=1}^n 
\end{equation}
where the \emph{index of disperson} $d=1$ when observations are independent or $d>1$ when nearby samples are correlated; the value of $d$ can be estimated from the data.

@@

\subsection{Probabilistic framework}
\TODO{This section is not strictly necessary. It helps relate probabilistic models to our approach, which does not rely on an explicity probabilistic model.}
Let $x_i \in \mathcal X, i=1\ldots,x_n$ denote a sample of consequtive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is the real-valued firing rate \TODO{Need to define rate more precisely}, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
We assume that observations are drawn from some true cumulative probability distribution $F: \mathcal X \mapsto R$. Then let $\hat F_n$ denote the empirical distribution function obtained from the observed sample, which itself is a random variable.
\begin{equation}
\hat F_n(x) = \frac 1 n \sum\limits_{i=1}^n \mathbf{1}(x \ge x_i)
\end{equation}
where $\mathbf 1(x \ge x_i)$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x_i$ and 0 otherwise. \TODO{KJ: Are you using cumulative distributions to avoid binning? DY: Yes. Many theorems such as Glivenko-Catelli Theorem and Skorohod's Representation are proven using  cumulative representation. The empirical distribution is nearly always represented in its cumulative form. Expressing convergence is requires binning or using cumulative; the latter is simpler.}

The observations need not be independently and identically sampled from $F$. The only assumption is that the data generating process is ergodic, \emph{i.e.}\;that $\max\limits_x \left[\hat F_n(x) - F(x)\right] \to 0$ as the sample size $n$ increases. Then each $x_i$ can be thought to be drawn from $F$ when not conditioned on nearby observations.

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of the true probability distribution
\begin{equation}
\Sigma = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif F(z)
\quad\mbox{where}\;
\bar x = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The \emph{plug-in estimator} $\hat\Sigma$ is obtained by plugging the empirical distribution $F_n$ in place of the true distribution $F$
\begin{equation}
\begin{split}
\hat\Sigma = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif \hat F_n(z) 
\equiv \frac  1 n\sum\limits_{i=1}^n (x_i-\bar x)(x_i - \bar x)
\\
\mbox{where}\;\bar x = 
\int\limits_{z\in \mathcal X}z \dif \hat F_n(z) \equiv \frac 1 n\sum\limits_{i=1}^n x_i
\end{split}
\end{equation}
The plug-in estimator is known to be biased toward zero because the estimate of the mean $\bar x$ absorbs a fraction of the variances.

\subsection{Measures of error}
Let $\hat\Sigma_\gamma$ be a family of approximate (biased) covariance estimates indexed by $\gamma$. To evaluate each estimate, let's define the \emph{loss function} $\mathcal L : \Theta \times \Theta \mapsto \mathbb R$ such that $\mathcal L(\hat S,S)$ is minimized when $\hat S = S$.  Then \emph{excess loss} $\ell(\hat S, S) = \mathcal L(\hat S, S) - \mathcal L(S,S)$ assumes zero at the minimum. \TODO{This is too abstract -- need to motivate with some examples maybe}

When the loss function is the mean squared error, which, for matrices, is called the normalized, squared Frobenius norm: 
\begin{equation}
\mathcal L_e(\hat S,S) =\tfrac 1 n \|\hat S-S\|_F^2 = \tfrac 1 n \Tr\left((\hat S-S)(\hat S-S)^\T\right)
\end{equation}
the variance-bias decomposition is the simple sum
\begin{equation}
r_\gamma =  b_\gamma^2 + \varepsilon_\gamma^2
\end{equation}

@@

An unbiased estimate does not guarantee lack of bias in nonlinear functions of that estimate: $\mathbb E\left[\hat\Sigma_0\right]=\Sigma \centernot\implies  E\left[\varphi(\hat\Sigma_0)\right]=\varphi(\Sigma)$, where $\varphi$ is a nonlinear function. For example, if $u_{\max}(S)$ is the largest eigevalue of square matrix $S$, then $\mathbb E\left[u_{\max}(\hat\Sigma_0)\right] > u_{\max}(\Sigma)$. Some regularization schemes have been designed to correct for the eigenspectrum bias  at the cost of adding bias to covariance coefficients \citep{Ledoit:2004}.

\paragraph{Evaluation of estimators:}
Another popular loss  function for covariance estimation arises from the theory of normal distributions.  Under the assumption of normality  with zero mean, the log likelihood of the covariance matrix $\Sigma$ given the independent sample $z_1,\ldots,z_m$, with $S = \tfrac 1 m \sum\limits_{i=1}^m z_i z_i^\T$,  
\begin{equation}
L\left(\Sigma \mid S\right) = -\tfrac m 2 \ln(2\pi) - \tfrac m 2 \ln \det \Sigma - \tfrac m 2 \Tr(\Sigma^{-1} S)
\end{equation}
Then the Gaussian loss function $\mathcal L_g$ is constructed by rescaling the normal log likelihood and dropping the constant term
\begin{equation}
\begin{split}
\mathcal L_g(\hat S,S) = &-\frac 2 m L\left(S \mid \hat S \right) - \ln(2\pi) 
\\ 
\equiv & \ln \det \hat S + \Tr(\hat S^{-1} S) 
\end{split}
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat S,S) = \mathcal L_g(\hat S,S) - \mathcal L_g(S,S) 
= -\ln \det \left(\hat S^{-1} S\right) + \Tr\left(\hat S^{-1}S\right) - p
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

>>>>

\subsection{Basic concepts}
Let $x_i \in \mathcal X, i=1\ldots,n$ denote a sample of consecutive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is denoted by the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
For example, in our applicaiton, the firing rate is estimated from somatic calcium fluorescence signals with the mean stimulus response subtracted, making it possible for elements of $x_i$ to take on negative values.

We do not assume that the observations $x_i$ are independently distributed. Rather, the data generating process is assumed to be \emph{ergodic}, \emph{i.e.}\;described by a \emph{true distirbution} $F: \mathcal X \mapsto  \mathbb [0, 1]$ (cumulative) over long periods of time such that the \emph{empirical distribution} $\hat F_n$ from a given sample will convege to $F$ with increasing sample size $n$.

Formally, the emprical distribution is defined as \TODO{this may be unecessary, but I leave it for now.}
\begin{equation}
\hat F_n(x) = \frac 1 n \sum\limits_{i=1}^n \mathbf{1}(x \ge x_i)
\end{equation}
where $\mathbf 1(x \ge x_i)$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x_i$ and 0 otherwise. \TODO{KJ: Are you using cumulative distributions to avoid binning? DY: Yes. Many theorems such as Glivenko-Catelli Theorem and Skorohod's Representation are proven using  cumulative representation. The empirical distribution is nearly always represented in its cumulative form. Expressing convergence is requires binning or using cumulative; the latter is simpler.}

And, for an ergodic process, $\max\limits_{x\in\mathcal X} \left|\hat F_n(x) - F(x)\right| \to 0$ as $n$ increases.

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of $F$:
\begin{equation}
\Sigma = \int\limits_{z\in\mathcal X} (z - \mu)(z - \mu)^\T \dif F(z)
\quad\mbox{where}\;
\mu = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The domain $\Theta$ is the set of all positive-definite $p\times p$ matrices, which is a cone in $\mathbb R^{p\times(p+1)/2}$.

But the usual estimator of the true covariance matrix $\Sigma$ is  the sample covariance matrix:
\begin{equation}
\hat\Sigma_0 =  \frac 1 {n-c} \sum\limits_{i=1}^n (x_i -\hat\mu) (x_i - \hat\mu)^\T
\quad\mbox{where}\;
\hat\mu = \frac 1 n \sum\limits_{i=1}^n x_i
\end{equation}
where $c$ is the sum of temporal correlations \TODO{there must be a better description of this}. For independent observations, $c=1$;  $c>1$ when nearby samples are correlated; the value of $c$ can be estimated from the data.

$\hat\Sigma_0$ is designed to be unbiased such that $\mathbb E\left[\hat\Sigma_0\right]=\Sigma$.
Here and below, $\mathbb E[\cdot]$ denotes the expected value under the  unknown true distribution $F$. All variables with a hat (\emph{e.g.}\;$\hat \Sigma_0,\hat \mu$) are functions of the empirical distribution $\hat F_n$, which itself is a random variable.

Nonlinear functions of an unbiased estimate can be biased: $\mathbb E\left[\hat\Sigma_0\right]=\Sigma \centernot\implies  E\left[\varphi(\hat\Sigma_0)\right]=\varphi(\Sigma)$.  For example, if $u_{\max}(S)$ is the largest eigevalue of square matrix $S$, then $\mathbb E\left[u_{\max}(\hat\Sigma_0)\right] > u_{\max}(\Sigma)$. Some estimators have been designed to correct for the eigenspectrum bias at the cost of adding bias to covariance coefficients \citep{Ledoit:2004}.

\subsection{Loss function}
The optimization of a covariance matrix estimate $\hat\Sigma$ is performed with respect to a \emph{loss function} $\mathcal L(\hat\Sigma,\Sigma)$, which expresses the discrepancy \TODO{right word?} between $\hat\Sigma$ and $\Sigma$ and attains its minimum when $\hat\Sigma=\Sigma$.  
Then \emph{excess loss}  
\begin{equation}
\ell(\hat\Sigma,\Sigma) = \mathcal L(\hat\Sigma,\Sigma)-\mathcal L(\Sigma,\Sigma)
\end{equation}
assumes zero at its minimum.

A particularly useful loss function is the mean squared error (MSE), which is proportional to the square of the Frobenius  norm $\|\cdot\|_F$ of the difference between the matrices: 
\begin{equation}
\mathcal L_e(\hat\Sigma,\Sigma) =\frac 1 p \|\hat\Sigma-\Sigma\|_F^2 = \frac 1 p \Tr\left((\hat \Sigma-\Sigma)(\hat\Sigma-\Sigma)^\T\right)
\end{equation}
The MSE is its own excess loss: $\ell_e(\hat\Sigma,\Sigma) \equiv \mathcal L_e(\hat\Sigma,\Sigma)$.

The Gaussian loss function $\mathcal L_g$  arises from the theory of normal distributions. When observations are identically and independently distributed according to a multivariate normal distribution with zero means, the log likelihood of the covariance matrix $\Sigma$ with $\hat\Sigma = \frac 1 n \sum\limits_{i=1}^n x_i x_i^\T$ is  
\begin{equation}
L\left(\Sigma \mid \hat\Sigma\right) = -\frac n 2 \ln(2\pi) - \frac n 2 \ln \det \Sigma - \frac n 2 \Tr(\Sigma^{-1} \hat \Sigma)
\end{equation}
Then $\mathcal L_g$ is constructed by rescaling $L\left(\Sigma \mid \hat\Sigma\right)$ and dropping the constant term:
\begin{equation}
\mathcal L_g(\hat\Sigma,\Sigma) 
=  -\frac 2 {pn} L\left(\Sigma \mid \hat\Sigma \right) - \frac 1 p \ln(2\pi) 
\equiv  \frac 1 p\left(\ln \det \hat \Sigma + \Tr(\hat \Sigma^{-1}) \right) 
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat\Sigma,\Sigma) = \mathcal L_g(\hat\Sigma,\Sigma) - \mathcal L_g(\Sigma,\Sigma)  
= \frac 1 p \left(-\ln \det (\hat \Sigma^{-1} \Sigma) + \Tr(\hat \Sigma^{-1}\Sigma)\right) - 1
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

\subsection{Emprical loss}

\subsection{Variance/bias decomposition}
the variance-bias decomposition is the simple sum
\begin{equation}
r_\gamma =  b_\gamma^2 + \varepsilon_\gamma^2
\end{equation}

Let $\hat\Sigma_\gamma$ be a family of approximate (biased) covariance estimates indexed by $\gamma$. To evaluate each estimate, let's define the \emph{loss function} $\mathcal L : \Theta \times \Theta \mapsto \mathbb R$ such that $\mathcal L(\hat S,S)$ is minimized when $\hat S = S$.  Then \emph{excess loss} $\ell(\hat S, S) = \mathcal L(\hat S, S) - \mathcal L(S,S)$ assumes zero at the minimum. \TODO{This is too abstract -- need to motivate with some examples maybe}

@@

@@

\paragraph{Evaluation of estimators:}
